---
title: "Example contributed script - create encounter histories from Motus detection data"
author: "Amie MacDonald (amacdonald@birdscanada.org)"
date: "2020-06-23"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Arrange Motus detection data into encounter histories that can be used in mark-recapture modelling with this code. Transform the detection data into a dataframe where columns represent time intervals and each row is an individual. The entries in each cell represent whether an individual was detected by a Motus station during that time interval in a single-state or multi-state framework.

## Setup

Load packages:

```{r packages}
library(tidyverse)
library(lubridate)

# set timezone
Sys.setenv(TZ = "UTC")

```

### Load sample dataset

Use the sample dataset from project 176, restricted to just Semipalmated Sandpipers. This dataset is saved as an RDS file in same folder as this script.

```{r data}

df.alltags <- readRDS("sample_data.rds")

```

## Transform detection data into encounter histories

### Single-state framework

This example uses daily time intervals and restricts the study period to August and September 2015. Each Semipalmated Sandpiper receives a `1` for days it was detected by a Motus station and a `0` for days where it was not detected. In case there are days where no birds were detected, a fake tag is created that was detected everyday. Then the dataframe is pivoteda and the fake tag is removed.

```{r transform}

det.data <- df.alltags %>% 
  select(motusTagID, ts) %>% # keep only the tag ID and the timestamp of the detection
  mutate(ts = date(ts)) %>% # keep only the date of the timestamp, not the time of day
  mutate(detect = 1) %>% # create a new column called detect and fill with 1 for each detection
  filter(ts < "2015-10-01") %>% # remove all data in October and after
  distinct() # remove duplicates (birds detected multiple times in the same day)

min(det.data$ts) # find the earliest detection

ts <- seq(ymd("2015-08-03"), ymd("2015-09-30"), by = "days") # create a sequence of all dates from the first 
                                                             # detection to the end of September

fake.df <- data.frame(ts) %>% # create a fake tag that was detected (with a false code) on all days
  mutate(motusTagID = 99999) %>% 
  mutate(detect = 99999)

det.data.for.eh <- bind_rows(det.data, fake.df) %>% # add the fake tag to the real detection data
  arrange(ts)

enc.hist <- det.data.for.eh %>% # create the encounter history
  tidyr::pivot_wider(id_cols = motusTagID, names_from = ts, values_from = detect, # transform the data
                     values_fill = list(detect = 0)) %>%                          # fill all empty cells with 0
  filter(!motusTagID == 99999) %>% # remove fake tag
  arrange(motusTagID)

```

### Multi-state framework

Create encounter histories where each location receives a different number in the encounter history (detection at site 1 is `1`, detection at site 2 is `2`, and no detection remains `0`). The process is the same as above, except the Motus station name is kept and assigned a number. In this example only the two stations in James Bay are kept.

```{r transform_multistate}

det.data <- df.alltags %>% 
  select(motusTagID, ts, recvDeployName) %>% # keep tag ID, timestamp, and receiver name
  mutate(ts = date(ts)) %>% # keep date but not time
  filter(recvDeployName %in% c("Netitishi", "North Bluff")) %>% # keep only receivers in James Bay
  distinct() # remove duplicates

min(det.data$ts) # find first detection
max(det.data$ts) # find last detection

ts <- seq(ymd("2015-08-03"), ymd("2015-09-08"), by = "days") # create a sequence of all days between first and last
                                                             # detection

fake.df <- data.frame(ts) %>% # create the fake tag detected every day
  mutate(motusTagID = 99999) %>% 
  mutate(recvDeployName = "fake")

det.data.for.ms.eh <- bind_rows(det.data, fake.df) %>% # add fake tag to detection data
  arrange(ts) %>% 
  mutate(recvDeployName = str_replace(recvDeployName, "North Bluff", "1")) %>% # replace receiver names with numbers
  mutate(recvDeployName = str_replace(recvDeployName, "Netitishi", "2")) %>% 
  mutate(recvDeployName = str_replace(recvDeployName, "fake", "99999"))

det.data.for.ms.eh$recvDeployName <- as.numeric(det.data.for.ms.eh$recvDeployName) # format as numbers 

ms.enc.hist <- det.data.for.ms.eh %>% # create encounter histories
  tidyr::pivot_wider(id_cols = motusTagID, names_from = ts, values_from = recvDeployName,
                     values_fill = list(recvDeployName = 0)) %>% # fill 0 for days a bird wasn't detected
  filter(!motusTagID == 99999) %>% 
  arrange(motusTagID)

```


